Seabird assemblages are linked to the major western boundary current off eastern Australia

### 2024-03-06 ms_PROOCE/ms_PROOCE_revised-markup.Rmd
update README, update scripts-workflow figure, updated references inlcuding Daudt et al 2024 and 'corrected' refs for R pckgs, plus all source RMD files

date: "2023-11-02"

@@

* The East Australian Current (EAC) is experiencing an ongoing poleward intensification
* We described seabird assemblages based on the Region of Common Profiles (RCP) method
* Assemblages of seabird species were correlated with the EAC
* Our analysis can be replicated to assess possible changes in seabird distribution
* The RCP method works well at the macro spatial scale for highly mobile animals

@@

Seabird assemblages have been widely described and related to environmental characteristics [@ribic1997; @woehler2003; @hyrenbach2007]. Considering the hierarchical nature of oceanographic processes and their resulting features [@haury1978patterns], it is not surprising that the distribution of highly mobile seabirds are linked differently across temporal and spatial scales [@haury1978patterns; @hunt1987]. In the marine environment, large-scale oceanographic features (e.g. boundary currents and water masses) act similarly to terrestrial biomes (based on vegetation), and influence seabird assemblage structure---i.e. the occurrence of species and their relative numerical composition [@ribic1997; @hyrenbach2007; @commins2014]. In contrast, local abundance is related to meso- (10--100 km) and sub-mesoscale (1--10 km) processes that aggregate prey and therefore predators, such as fronts and eddies [@weimerskirch2007; @scales2014]. Such processes, although relatively ephemeral, are primarily driven by physical and atmospheric forces, and are thus somewhat predictable in space and time, providing consistent feeding areas for marine predators when they occur [@davoren2013; @hazen2013]. Sea surface temperatures are a key driver of seabird distributions, at individual- and assemblage-levels, operating over a range of spatial and temporal scales from macro (e.g. water masses) to fine (e.g. fronts) [@dunlop1988; @ribic1997; @quillfeldt2015]. Despite extensive surveys [e.g. @ballance1997] and tracking studies [e.g. @carneiro2020], many oceanic regions still lack information on the distributions and abundances of seabird species, and how they assemble in space and time.

@@

In light of eastern Australia's importance for seabirds, and the recent attention on how predicted climate change will influence the EAC's marine biota [@poloczanska2007], an assessment of the contemporary spatial distributions of seabirds within the EAC at large temporal and spatial scales is timely. Seabirds can serve as sentinels for tracking ecological changes in the EAC in many ways [e.g. @sydeman2012]. Surprisingly though, seabird distribution is under-represented in climate change studies, despite being highlighted as a priority topic [@grémillet2009; @sydeman2012].

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\~18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC [@kessler2013; @ganachaud2014]. The South Caledonian Jet completes the EAC system at \~22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (**Fig.** \@ref(fig:fig1-study-area)**a**; @kessler2013; @ganachaud2014). The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies [@oke2000; @ridgway2003a; @oke2019]. The EAC flows polewards and is primarily associated with the shelf break, but at \~32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea [@mata2006; @oke2019], the 'EAC eastern extension', often termed the 'Tasman Front' [@oke2019a]. Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the 'EAC southern extension'; **Fig.** \@ref(fig:fig1-study-area)**a**), carrying eddy-like features as a transient current [@oke2019].

@@

```{r fig1-study-area, fig.cap = "\\label{fig1-study-area} Study area showing the main ocean currents (a), and seabird records made by season on top of the 1° latitude × 1° longitude grid cells (b). In (a), the East Australian Current (EAC) system is highlighted with its acronyms in bold and drawn in blue. The South Equatorial Current (SEC) and its branches, the North Caledonian Jet (NCJ) and South Caledonian Jet (SCJ), are drawn in red. The EAC 'eastern extension' is an eddy field (blue shaded area) often called the 'Tasman Front', from which the East Auckland Current originates (EAUC, in green). Figure (a) is adapted after Oke, Roughan, et al. (2019); for colour references see the online version", fig.align = "center", out.width = "100%", echo = FALSE}

@@

Seabird data were gathered aboard *RV Investigator* and *RV Falkor* on 15 voyages, between 2016/17--2020/21, spanning over \~37° of latitude (**Fig.** \@ref(fig:fig1-study-area)**b**; **Supplementary Table 1**). One or more observers continuously recorded seabirds within a 300 m forward quadrant (eye height 25 m a.s.l) from sunrise to sunset hours while the ship was underway. The species, the numbers of birds, and their behaviours were logged in real time into a ship-based web server. Environmental parameters for atmospheric and oceanographic conditions were logged continuously by automated-ship based sensors, with seabird observations supplementing relevant environmental conditions such as visibility or associations with marine mammals. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species). Birds following the vessels were excluded from analyses. Seabird observation protocols followed @raymond2010, in use around Australia since 1980.

@@

```{r table1, echo = FALSE}
data.frame(
  Variable = c("Sea Surface Temperature", "Sea Surface Temperature gradient", "Sea Surface Salinity", "Mixed Layer Depth", "8-day average Chlorophyll-a", "Log10 8-day average Chlorophyll-a", "Eddy Kinetic Energy", "Climatic mean EKE", "Climatic standard deviation EKE", "Climatic mean SSTgrad", "Bathymetry", "Slope", "Distance from coast"),
  # Abbreviation needs the `str_replace_all` because I'm using `escape=F` in `kbl()`
  Abbreviation = stringr::str_replace_all(c("sst", "sst_grad", "sss", "mld", "chl8", "log10_chl8", "eke", "clim_eke_mean", "clim_eke_sd", "clim_sst_grad", "bat", "slope", "dist_coast"), "_", "\\\\_"), 
  Units = c("°Celsius", paste0(parse(text = "\u0394"), "°Celsius"), "PSU", "m", "mg/m${^3}$", "$log_{10}$(chl8)", "m/s", "mean(m/s)/season/10 years", "sd(m/s)/season/10 years", paste0("mean(", paste0(parse(text = "\u0394"), "°Celsius"), ")/season/10 years"), "m", "degrees (°)", "km"),
  `Spatial resolution` = c("0.25° × 0.25°", "0.25° × 0.25°", "0.083° × 0.083°", "0.083° × 0.083°", "0.036° × 0.036°", "0.036° × 0.036°", "0.2° × 0.2°", "0.2° × 0.2°", "0.2° × 0.2°", "0.25° × 0.25°", "0.004° × 0.004°", "0.004° × 0.004°", "NA"),
  `Temporal resolution` = c("daily", "daily", "daily", "daily", "8-days composite", "8-days composite", "daily", "seasonal climatology", "seasonal climatology", "seasonal climatology", "NA", "NA", "NA"),
  Source = c("Huang et al. (2021)${^a}$", "Derived from Huang et al. (2021)${^a}$", "E.U. Copernicus Marine Service Information${^b}{^c}$", "E.U. Copernicus Marine Service Information${^b}{^c}$", "Hu et al. (2012)${^d}$", "Derived from Hu et al. (2012)${^d}$", "IMOS/AODN${^e}$", "Derived from IMOS/AODN${^e}$", "Derived from IMOS/AODN${^e}$", "Derived from Huang et al. (2021)${^a}$", "GEBCO${^f}$", "Derived from GEBCO${^f}$", "Calculated in R (see main text)")
) %>% 
  kableExtra::kbl(caption = "\\label{table1}Environmental data used in the Region of Common Profile models applied to seabird data off eastern Australia, their spatial/temporal resolutions, units, and sources", booktabs = TRUE, escape = FALSE)  %>% 
  kableExtra::row_spec(0, bold = TRUE) %>% 
  kableExtra::column_spec(1, width = "4cm") %>% 
  kableExtra::column_spec(6, width = "3cm") %>% 
  kableExtra::footnote(
    alphabet = c(
    "NOAA OISST: (https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html); ", 
    "2011--2019: GLOBAL_MULTIYEAR_PHY_001_030 (https://doi.org/10.48670/moi-00021); ", 
    "2020--2021: GLOBAL_ANALYSIS_FORECAST_PHY_001_024 (https://doi.org/10.48670/moi-00016); ",
    "ERDDAP/NOAA (https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla8day.html); ",
    "IMOS/AODN (https://portal.aodn.org.au/search), 'OceanCurrent - Gridded sea level anomaly - Delayed mode' data set; ",
    "GEBCO (https://www.gebco.net/data_and_products/gridded_bathymetry_data/)"))
```

@@

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods [@biology2002]. Therefore, seasonal differences in species composition were expected [e.g. @reid2023], and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling datasets for each season. We aggregated seabird data by 1° latitude $\times$ 1° longitude grid cells (*c.* 110 $\times$ 110 km) and season (**Fig.** \@ref(fig:fig1-study-area)**b**), by summing seabirds from the same species. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \>1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

The model is specified in terms of assemblages and site-specific probabilities. There are $K$ assemblages (i.e. RCPs), $K = (1, ..., k)$, that are common across the study area. The outcome variable is a vector $y_{i} = (y_{i1}, ... y_{iS})$, that gives the count or presence-absence of each of the $S$ species at site $i$. Depending on whether we model the counts or presence-absence data, the assemblages are defined in terms of the expected frequency or relative frequency of the species, respectively. An output of the model is the estimated probability that site $i$ has each of the $K$ possible assemblages. RCPs are a multivariate adaptation of a mixture-of-experts model; as such, the mixing proportions (assemblage structure) vary according to covariates (predictors) [@foster2013]. RCPs can be directly interpreted as assemblages; the assemblage structure is given by the species profiles.

@@

A total of 10,261 seabird records was obtained during the 15 voyages, comprising 142,646 individuals from 80 species, belonging to 11 families and five orders. The five most abundant species made up 83% of total birds counted (43.0% short-tailed shearwater *Ardenna tenuirostris*, 17.2% fairy prion *Pachyptila turtur*, 14.3% wedge-tailed shearwater *Ardenna pacifica*, 5.2% sooty tern *Onychoprion fuscatus*, 3.2% shy albatross *Thalassarche cauta*). The five most frequently encountered species were *A. pacifica* (frequency of occurrence = 35.3%), *T. cauta* (33.2%), brown booby *Sula leucogaster* (25.4%), red-footed booby *S. sula* (25.4%) and black-browed albatross *T. melanophris* (25%) (**Supplementary Fig. 1**). An increase in species richness towards the south of the study region was clear in all but the winter season (**Supplementary Fig. 2**). Noticeably higher counts were made around Tasmania, possibly due to the 'colony effect' and/or the presence of Southern Ocean species (**Supplementary Fig. 2**). However, in summer, one cell in the Coral Sea had counts of the same order of magnitude as off Tasmania.

@@

For all models, two RCPs was the optimal number of seabird groups, except for the presence-absence-autumn model, which indicated three RCPs as the best fit (**Supplementary Figs. 3, 4**). The best fitting models generally had one or two covariates selected, although the abundance-autumn model had five covariates (**Table** \@ref(tab:table2)). The best models showed a good fit with no apparent deviations from the assumptions (**Supplementary Figs. 5, 6**). All the best models selected sea surface temperature ('sst') or salinity ('sss'), as essential for characterising water masses. The thresholds for defining RCP membership were 18--21°C for 'sst' and 35.1 PSU for 'sss' (**Fig.** \@ref(fig:fig2-partial-plots), **Supplementary Figs. 7, 8**), similar to the characteristics used to to define the EAC. Furthermore, most of the best models selected bathymetry ('bat'), suggesting a gradient between 'coastal' and 'oceanic' taxa (**Supplementary Figs. 7, 8**). Of particular note, proxies for persistent oceanographic features that may aggregate food resources (the 'climatic' variables) for seabirds were selected only in the autumn and spring abundance models.

@@

```{r fig2-partial-plots, fig.cap = "\\label{fig2-partial-plots}Partial plots for the retained covariates in the best spring (a) presence-absence and (b) abundance models. The plot shows the fitted probability of belonging to a Region of Common Profiles (RCP) against the environmental value. All other predictors are held at their mean values. Results for other seasons and models can be found in the \\textbf{Supplementary Figs. 7, 8}. Refer to Table 1 for the environmental data acronyms", fig.align = "center", out.width = "100%", echo = FALSE}

@@

Grid point-predictions are the assignment of the RCP with the highest probability value for a given cell. When predicting RCP onto space, patterns of point-predictions suggest a biogeographic transition in seabird species around \~34°S, with both presence-absence and abundance models pointing to similar results (**Fig.** \@ref(fig:fig3-point-pred)). These patterns are further reinforced when looking at the mean probability along with the lower and upper confidence intervals of predictions (**Fig.** \@ref(fig:fig4-prob-maps), **Supplementary Figs. 9, 10**), showing that around \~34°S, the 'northern' RCP tended to have higher probabilities in coastal regions, whereas the 'southern' RCP spread northward offshore. The autumn presence-absence model (three RCPs) had a similar broad-scale pattern, although the southern RCP is farther south, and a strong 'Tasmania/Bass Strait' assemblage appears (**Fig.** \@ref(fig:fig3-point-pred), **Supplementary Figs. 9, 10**).

```{r fig3-point-pred, fig.cap = "\\label{fig3-point-pred}Spatialized point predictions (hard-classes) for seabird assemblages (Region of Common Profile; RCP), by season, based on presence-absence (a) and abundance (b) models, off eastern Australia. Each grid was assigned to the RCP group with highest membership probability", fig.align = "center", out.width = "110%", echo = FALSE}

@@

```{r fig4-prob-maps, fig.cap = "\\label{fig4-prob-maps}Predicted probability membership of each seabird assemblage (Region of Common Profiles; RCP) and grid, off eastern Australia, from spring presence-absence (a) and abundance (b) models. The central column, 'mean', corresponds to the point prediction and Bayesian boot-strapped, lower and upper confidence intervals (CI), on its sides. Note that, in case of two RCPs, RCP2 probability equals (1 - RCP1 probability). Results for other seasons and models can be found in the \\textbf{Supplementary Figs. 9, 10}", fig.align = "center", out.width = "110%", echo = FALSE}

@@

The broad pattern of two large-scale assemblages is reflected in their species compositions. Although there are overlaps among the seabird groups, both presence-absence and abundance models captured typical tropical species in the northern RCP (e.g. boobies *Sula* spp., lesser frigatebird *Fregata ariel*) and typical Southern Ocean species in the southern RCP (e.g. small albatrosses *Thalassarche* spp., fairy prion *P. turtur*) (**Fig.** \@ref(fig:fig5-spp-profiles), **Supplementary Figs. 11, 12**). The autumn presence-absence model, which had three RCPs, also showed the characteristic taxa representing the 'northern' and 'southern' RCPs. However, the intermediate 'Tasmania/Bass Strait' assemblage showed a mixture of both groups with shy albatross and Australasian gannet *Morus serrator* prominent with high probabilities of occurrence.

@@

```{r fig5-spp-profiles, fig.cap = "\\label{fig5-spp-profiles}Species profiles of seabird assemblages (Region of Common Profile; RCP) off eastern Australia in spring. Values are the average and 95\\% confidence intervals of probability of occurrence (a) and predicted mean abundance (b) for each species, based on 1000 Bayesian bootstraps. Values in (b) were log10-transformed to accommodate the high variation between species. Results for other seasons and models can be found in the \\textbf{Supplementary Figs. 11, 12}", fig.align = "center", out.width = "80%", echo = FALSE}

@@

To the best of our knowledge, this is the first study applying RCPs to highly mobile animals such as seabirds. RCP models have been previously applied to plants [@lyons2017], and relatively sedentary animals, such as macrobenthic organisms [@pantiukhin2021] and demersal fish [@hill2017]. Oceanic seabirds can have daily movements of hundreds of kilometres [e.g. @mcduie2015], during which they traverse a variety of habitats and environmental gradients. As such, capturing the distribution of highly mobile species, and their relationships with environmental parameters is not a trivial task, and much harder than for sessile organisms.

>>>>

date: "`r Sys.Date()`"

@@

-   \textcolor{red}{\sout{The East Australian Current (EAC) is experiencing an ongoing poleward intensification}}

-   We described seabird assemblages based on the Region of Common Profiles (RCP) method

-   Assemblages of seabird species were correlated with the \textcolor{red}{East Australian Current}

-   Our analysis can be replicated to assess possible changes in seabird distribution

-   The RCP method works well at the macro spatial scale for highly mobile animals

@@

Seabird assemblages have been widely described and related to environmental characteristics [@ribic1997; @woehler2003; @hyrenbach2007]. Considering the hierarchical nature of oceanographic processes and their resulting features [@haury1978patterns], it is not surprising that the distribution of highly mobile seabirds are linked differently across temporal and spatial scales [@haury1978patterns; @hunt1987]. In the marine environment, large-scale oceanographic features (e.g. boundary currents and water masses) act similarly to terrestrial biomes (based on vegetation), and influence seabird assemblage structure---i.e. the occurrence of species and their relative numerical composition [@ribic1997; @hyrenbach2007; @commins2014]. In contrast, local abundance is related to meso- (10--100 km) and sub-mesoscale (1--10 km) processes that aggregate prey and therefore predators, such as fronts and eddies [@weimerskirch2007; @scales2014]. Such processes, although relatively ephemeral, are primarily driven by physical and atmospheric forces, and are thus somewhat predictable in space and time, providing consistent feeding areas for marine predators when they occur [@davoren2013; @hazen2013]. Sea surface temperatures are a key driver of seabird distributions, at individual- and assemblage-levels, operating over a range of spatial and temporal scales from macro (e.g. water masses) to fine (e.g. fronts) [@dunlop1988; @ribic1997; @quillfeldt2015; @gall2022; @serratosa2020]. Despite extensive surveys [e.g. @ballance1997] and tracking studies [e.g. @carneiro2020], many oceanic regions still lack information on the distributions and abundances of seabird species, and how they assemble in space and time.

@@

In light of eastern Australia's importance for seabirds, and the recent attention on how predicted climate change will influence the EAC's marine biota [@poloczanska2007], an assessment of the contemporary spatial distributions of seabirds within the EAC at large temporal and spatial scales is timely. Seabirds can serve as sentinels for tracking ecological changes in the EAC in many ways [e.g. @sydeman2012]. Surprisingly though, \textcolor{red}{particularly in the Australasian region,} seabird distribution is under-represented in climate change studies, despite being highlighted as a priority topic [@grémillet2009; @sydeman2012].

@@

The EAC originates from the South Equatorial Current (SEC), a tropical nutrient-poor current. When the SEC's North Caledonian Jet meets the Queensland Plateau (\~18°S), it splits into a north-flowing branch towards the Gulf of Papua and a south-flowing branch, which marks the start of the EAC [@kessler2013; @ganachaud2014]. The South Caledonian Jet completes the EAC system at \~22°S after encountering the shelf break in the southern portion of the Great Barrier Reef (**Fig.** \@ref(fig:fig-study-area)**a**; @kessler2013; @ganachaud2014). The EAC interacts with the bathymetry of the seafloor, giving rise to oceanographic features and processes, such as smaller currents, upwellings, and persistent dynamic eddies [@oke2000; @ridgway2003a; @oke2019]. The EAC flows polewards and is primarily associated with the shelf break, but at \~32°S it detaches from the shelf and spawns an area of eddy generation that dissipates throughout the western Tasman Sea [@mata2006; @oke2019], the 'EAC eastern extension', often termed the 'Tasman Front' [@oke2019a]. Although most EAC energy is converted into eddies, the EAC still runs southwards following the shelf break towards Tasmania (the 'EAC southern extension'; **Fig.** \@ref(fig:fig-study-area)**a**), carrying eddy-like features as a transient current [@oke2019].

@@

```{r fig-study-area, fig.cap = "\\label{fig-study-area} Study area showing the main ocean currents (a), and seabird records made by season on top of the 1° latitude × 1° longitude grid cells (b). In (a), the East Australian Current (EAC) system is highlighted with its acronyms in bold and drawn in blue. The South Equatorial Current (SEC) and its branches, the North Caledonian Jet (NCJ) and South Caledonian Jet (SCJ), are drawn in red. The EAC 'eastern extension' is an eddy field (blue shaded area) often called the 'Tasman Front', from which the East Auckland Current originates (EAUC, in green). Figure (a) is adapted after Oke, Roughan, et al. (2019); for colour references see the online version", fig.align = "center", out.width = "100%", echo = FALSE}

@@

Seabird data were gathered aboard *RV Investigator* and *RV Falkor* on 15 voyages, between 2016/17--2020/21, spanning over \~37° of latitude (**Fig.** \@ref(fig:fig-study-area)**b**; **Supplementary Table 1**). \textcolor{red}{Two} observers continuously recorded seabirds within a 300 m forward quadrant (eye height 25 m a.s.l) from sunrise to sunset hours while the ship was underway\textcolor{red}{, following a strip-transect protocol}. The species, the numbers of birds, and their behaviours \textcolor{red}{(i.e. flying, feeding, sitting on the water)} were logged in real time into a ship-based web server. \textcolor{red}{Seabird surveys were halted when sea conditions exceeded 6 m wave height and 40 kn of wind or visibility was under 500 m.} Environmental parameters for atmospheric and oceanographic conditions were logged continuously by automated-ship based sensors, with seabird observations supplementing relevant environmental conditions such as visibility or associations with marine mammals. Individual records received a date/time/geographic coordinate label (multi-species records have the same label but are logged as separate entries for each species). Birds following the vessels were excluded from analyses. Seabird observation protocols followed @raymond2010, in use around Australia since 1980.

@@

```{r table1, echo = FALSE}
data.frame(
  Variable = c("Sea Surface Temperature", "Sea Surface Temperature gradient", "Sea Surface Salinity", "Mixed Layer Depth", "8-day average Chlorophyll-a", "Log10 8-day average Chlorophyll-a", "Eddy Kinetic Energy", "Climatic mean EKE", "Climatic standard deviation EKE", "Climatic mean SSTgrad", "Bathymetry", "Slope", "Distance from coast"),
  # Abbreviation needs the `str_replace_all` because I'm using `escape=F` in `kbl()`
  Abbreviation = stringr::str_replace_all(c("sst", "sst_grad", "sss", "mld", "chl8", "log10_chl8", "eke", "clim_eke_mean", "clim_eke_sd", "clim_sst_grad", "bat", "slope", "dist_coast"), "_", "\\\\_"), 
  Units = c("°Celsius", paste0(parse(text = "\u0394"), "°Celsius"), "PSU", "m", "mg/m${^3}$", "$log_{10}$(chl8)", "m/s", "mean(m/s)/season/11 years", "sd(m/s)/season/11 years", paste0("mean(", paste0(parse(text = "\u0394"), "°Celsius"), ")/season/11 years"), "m", "degrees (°)", "km"),
  `Spatial resolution` = c("0.25° × 0.25°", "0.25° × 0.25°", "0.083° × 0.083°", "0.083° × 0.083°", "0.036° × 0.036°", "0.036° × 0.036°", "0.2° × 0.2°", "0.2° × 0.2°", "0.2° × 0.2°", "0.25° × 0.25°", "0.004° × 0.004°", "0.004° × 0.004°", "NA"),
  `Temporal resolution` = c("daily", "daily", "daily", "daily", "8-days composite", "8-days composite", "daily", "seasonal climatology", "seasonal climatology", "seasonal climatology", "NA", "NA", "NA"),
  Source = c("Huang et al. (2021)${^a}$", "Derived from Huang et al. (2021)${^a}$", "E.U. Copernicus Marine Service Information${^b}{^c}$", "E.U. Copernicus Marine Service Information${^b}{^c}$", "Hu et al. (2012)${^d}$", "Derived from Hu et al. (2012)${^d}$", "IMOS/AODN${^e}$", "Derived from IMOS/AODN${^e}$", "Derived from IMOS/AODN${^e}$", "Derived from Huang et al. (2021)${^a}$", "GEBCO${^f}$", "Derived from GEBCO${^f}$", "Calculated in R (see main text)")
) %>% 
  kableExtra::kbl(caption = "\\label{table1}Environmental data used in the Region of Common Profile models applied to seabird data off eastern Australia, their spatial/temporal resolutions, units, and sources", booktabs = TRUE, escape = FALSE)  %>% 
  kableExtra::row_spec(0, bold = TRUE) %>% 
  kableExtra::column_spec(1, width = "4cm") %>% 
  kableExtra::column_spec(6, width = "3cm") %>% 
  kableExtra::footnote(
    alphabet = c(
    "NOAA OISST: (https://psl.noaa.gov/data/gridded/data.noaa.oisst.v2.highres.html); ", 
    "2011--2019: GLOBAL_MULTIYEAR_PHY_001_030 (https://doi.org/10.48670/moi-00021); ", 
    "2020--2021: GLOBAL_ANALYSIS_FORECAST_PHY_001_024 (https://doi.org/10.48670/moi-00016); ",
    "ERDDAP/NOAA (https://coastwatch.pfeg.noaa.gov/erddap/griddap/erdMH1chla8day.html); ",
    "IMOS/AODN (https://portal.aodn.org.au/search), 'OceanCurrent - Gridded sea level anomaly - Delayed mode' data set; ",
    "GEBCO (https://www.gebco.net/data_and_products/gridded_bathymetry_data/)"))
```

@@

Seabirds are highly mobile animals, and many species migrate and/or disperse during their non-breeding periods [@biology2002]. Therefore, seasonal differences in species composition were expected [e.g. @reid2023], and in the probabilities of occurrence and in total numbers. Thus, we built separate modelling datasets for each season. \textcolor{red}{Seasons were defined as summer (Dec--Feb), autumn (Mar--May), winter (Jun--Aug), and spring (Sep--Nov).} We aggregated seabird data by 1° latitude $\times$ 1° longitude grid cells (*c.* 110 $\times$ 110 km) and season (**Fig.** \@ref(fig:fig-study-area)**b**), by summing seabirds from the same species \textcolor{red}{(flying, feeding, and sitting on the water)}. To test if the presence-absence and abundance (raw counts) data would result in similar assemblage patterns, we built two species matrices, where the presence-absence matrix transformed values \>1 to 1. To avoid numerical instabilities in the models prompted by unusual or rare species, we kept only species occurring in at least six grids within each season (summer = 12 species; autumn = 21; winter = 20; spring = 33). The associated environmental data of each seabird record were averaged within the same grid cells and seasons, and environmental data were standardised before modelling.

@@

The model is specified in terms of assemblages and site-specific probabilities. There are $K$ assemblages (i.e. RCPs), $K = (1, ..., k)$, that are common across the study area. The outcome variable is a vector $y_{i} = (y_{i1}, ... y_{iS})$, that gives the count or presence-absence of each of the $S$ species at site $i$. Depending on whether we model the counts or presence-absence data, the assemblages are defined in terms of the expected frequency or relative frequency of the species, respectively. An output of the model is the estimated probability that site $i$ has each of the $K$ possible assemblages. RCPs are a multivariate adaptation of a mixture-of-experts model; as such, the mixing proportions (assemblage structure) vary according to covariates (predictors) [@foster2013]. RCPs can be directly interpreted as assemblages; the assemblage structure is given by the species profiles \textcolor{red}{(i.e. the proportion of each species within the assemblage given by their probability of occurrence or predicted abundance)}.

@@

A total of 10,261 seabird records was obtained during the 15 voyages, comprising 142,646 individuals from 80 species, belonging to 11 families and five orders. The five most abundant species made up 83% of total birds counted (43.0% short-tailed shearwater *Ardenna tenuirostris*, 17.2% fairy prion *Pachyptila turtur*, 14.3% wedge-tailed shearwater *Ardenna pacifica*, 5.2% sooty tern *Onychoprion fuscatus*, 3.2% shy albatross *Thalassarche cauta*). The five most frequently encountered species were *A. pacifica* (frequency of occurrence = 35.3%), *T. cauta* (33.2%), brown booby *Sula leucogaster* (25.4%), red-footed booby *S. sula* (25.4%) and black-browed albatross *T. melanophris* (25%) (**Supplementary Fig. 1**). An increase in species richness towards the south of the study region was clear in all but the winter season (**Fig.** \@ref(fig:fig-spprich-nbirds)). Noticeably higher counts were made around Tasmania, possibly due to the 'colony effect' and/or the presence of Southern Ocean species (**Fig.** \@ref(fig:fig-spprich-nbirds)). However, in summer, one cell in the Coral Sea had counts of the same order of magnitude as off Tasmania.

```{r fig-spprich-nbirds, fig.cap = "\\label{fig-spprich-nbirds}Species richness and the total number of seabirds counted off eastern Australia during Australasian Seabird Group's ship-based surveys between 2016--2021, by grid/season, after data aggregation (see Methods). Dots represent the grid centroids.", fig.align = "center", out.width = "100%", echo = FALSE}

knitr::include_graphics("../results/FigS2_map-effort-spp-richness-and-total-birds-seasons.png")
```

@@

\textcolor{red}{In this section, we will present only the main results for the spring season when delving into specifics about each model and its results. The results of spring models summarise the key findings of our analysis, but we encourage the reader to look at the Supplementary Material for other seasons.}

For all models, two RCPs was the optimal number of seabird groups, except for the presence-absence-autumn model, which indicated three RCPs as the best fit (**Supplementary Figs. 2, 3**). The best fitting models generally had one or two covariates selected, although the abundance-autumn model had five covariates (**Table** \@ref(tab:table2)). The best models showed a good fit with no apparent deviations from the assumptions (**Supplementary Figs. 4, 5**). All the best models selected sea surface temperature ('sst') or salinity ('sss'), as essential for characterising water masses. The thresholds for defining RCP membership were 18--21°C for 'sst' and 35.1 PSU for 'sss' (**Fig.** \@ref(fig:fig-partial-plots), **Supplementary Figs. 6, 7**), similar to the characteristics used to to define the EAC. Furthermore, most of the best models selected bathymetry ('bat'), suggesting a gradient between 'coastal' and 'oceanic' taxa (**Supplementary Figs. 6, 7**). Of particular note, proxies for persistent oceanographic features that may aggregate food resources (the 'climatic' variables) for seabirds were selected only in the autumn and spring abundance models.

@@

```{r fig-partial-plots, fig.cap = "\\label{fig-partial-plots}Partial plots for the retained covariates in the best spring (a) presence-absence and (b) abundance models. The plot shows the fitted probability of belonging to a Region of Common Profiles (RCP) against the environmental value. All other predictors are held at their mean values. Results for other seasons and models can be found in the \\textbf{Supplementary Figs. 6, 7}. Refer to Table 1 for the environmental data acronyms", fig.align = "center", out.width = "100%", echo = FALSE}

@@

Grid point-predictions are the assignment of the RCP with the highest probability value for a given cell. When predicting RCP onto space, patterns of point-predictions suggest a biogeographic transition in seabird species around \~34°S, with both presence-absence and abundance models pointing to similar results (**Fig.** \@ref(fig:fig-point-pred)). These patterns are further reinforced when looking at the mean probability along with the lower and upper confidence intervals of predictions (**Fig.** \@ref(fig:fig-prob-maps), **Supplementary Figs. 8, 9**), showing that around \~34°S, the 'northern' RCP tended to have higher probabilities in coastal regions, whereas the 'southern' RCP spread northward offshore. The autumn presence-absence model (three RCPs) had a similar broad-scale pattern, although the southern RCP is farther south, and a strong 'Tasmania/Bass Strait' assemblage appears (**Fig.** \@ref(fig:fig-point-pred), **Supplementary Figs. 8, 9**).

```{r fig-point-pred, fig.cap = "\\label{fig-point-pred}Spatialized point predictions (hard-classes) for seabird assemblages (Region of Common Profile; RCP), by season, based on presence-absence (a) and abundance (b) models, off eastern Australia. Each grid was assigned to the RCP group with highest membership probability", fig.align = "center", out.width = "110%", echo = FALSE}

@@

```{r fig-prob-maps, fig.cap = "\\label{fig-prob-maps}Predicted probability membership of each seabird assemblage (Region of Common Profiles; RCP) and grid, off eastern Australia, from spring presence-absence (a) and abundance (b) models. The central column, 'mean', corresponds to the point prediction and Bayesian boot-strapped, lower and upper confidence intervals (CI), on its sides. Note that, in case of two RCPs, RCP2 probability equals (1 - RCP1 probability). Results for other seasons and models can be found in the \\textbf{Supplementary Figs. 8, 9}", fig.align = "center", out.width = "110%", echo = FALSE}

@@

The broad pattern of two large-scale assemblages is reflected in their species compositions. Although there are overlaps among the seabird groups, both presence-absence and abundance models captured typical tropical species in the northern RCP (e.g. boobies *Sula* spp., lesser frigatebird *Fregata ariel*) and typical Southern Ocean species in the southern RCP (e.g. small albatrosses *Thalassarche* spp., fairy prion *P. turtur*) (**Fig.** \@ref(fig:fig-spp-profiles), **Supplementary Figs. 10, 11**). The autumn presence-absence model, which had three RCPs, also showed the characteristic taxa representing the 'northern' and 'southern' RCPs. However, the intermediate 'Tasmania/Bass Strait' assemblage showed a mixture of both groups with shy albatross and Australasian gannet *Morus serrator* prominent with high probabilities of occurrence.

@@

```{r fig-spp-profiles, fig.cap = "\\label{fig-spp-profiles}Species profiles of seabird assemblages (Region of Common Profile; RCP) off eastern Australia in spring. Values are the average and 95\\% confidence intervals of probability of occurrence (a) and predicted mean abundance (b) for each species, based on 1000 Bayesian bootstraps. Values in (b) were log10-transformed to accommodate the high variation between species. Results for other seasons and models can be found in the \\textbf{Supplementary Figs. 10, 11}. Note the species on the Y-axis are ordered alphabetically, from bottom up.", fig.align = "center", out.width = "80%", echo = FALSE}

@@

To the best of our knowledge, this is the first study applying RCPs to highly mobile animals such as seabirds. RCP models have been previously applied to plants [@lyons2017], and relatively sedentary animals, such as macrobenthic organisms [@pantiukhin2021] and demersal fish [@hill2017]. Oceanic seabirds can have daily movements of hundreds of kilometres [e.g. @mcduie2015], during which they traverse a variety of habitats and environmental gradients. As such, capturing the distribution of highly mobile species, and their relationships with environmental parameters is not a trivial task, and much harder than for sessile organisms. \textcolor{red}{Among marine megafauna, seabirds are relatively easy to detect visually, enabling suitable sample sizes to be gathered over large areas} [@ballance2007]\textcolor{red}{. Thus, seabirds can be used as models to study the distribution of highly mobile animals in relation to environmental characteristics in the marine environment.}
